Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-12-01 manuscript.Rmd
update 2022-12-01 08:01:21

There are several use cases for which LCA methods are suitable. 
Although latent class analysis is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that latent class analysis is similar to confirmatory factor analysis,
but with a categorical latent variable, it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
One challenge is that, unlike confirmatory factor analysis,
LCA does not provide absolute fit indices that can be used to perform inference.
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values.
A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be utilized as a diagnostic
aid when using several clinical indicators to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental health
[@wu_abuse_2011] problems, or health outcomes after recovering from
cancer [@hardy_exploring_2018]. The example of shoe size is a rudimentary
illustration of this type of application. As another example, a recent study 
identified two suicide risk profiles in military service members and veterans
[@gromatsky_characteristics_2022].  One class was defined by higher-risk
history and another by lower-risk history. The higher-risk group comprised of
individuals with greater intent to die using methods with higher lethality,
and the lower-risk group was characterized by greater ambivalence towards death
and usage of lower lethality methods.

A related use case is where class-membership is known but imperfectly measured.
Mixture models can also be useful to evaluate the quality of measurement
across instruments. For example, a recent study had employment status 
information collected across register and survey data. There was disagreement 
between the two data sources. While it might be tempting to assume that the
register data is free of error, this has been shown to be false.
The authors used a mixture model to evaluate the underlying factor of 
"error corrected employment status" by using two indicators of the same question
and identifying random and systematic measurement error in both indicators 
[@pavlopoulos_measuring_2015].

LCA can be used as a descriptive analysis where a researcher wishes to
describe their sample and identify a few prototypes based on many
variables. As an example, if a survey among all Dutch academics was
carried out with the goal of bringing about a funding reform, LCA could
be used to discover the types of publications that get funding.
Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean.
Prior to this study, ways to categorize ocean microplastics in terms of length,
width and polymer type were rather arbitrary, and assumed that these properties
were independent. Also, there was a lack of information regarding variability 
within these categories. The authors used finite Gaussian mixture modeling to 
build a classification system for ocean microplastic particles based on a set 
of measured characteristics. With regards to particle dimensions, two classes
emerged: one with smaller fragments whose width and height were highly 
correlated, and another consisting of larger fragments with low correlation 
between particle width and height.

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group,
but do not know how many groups there are, LCA may be an appropriate
technique to answer this question. For example, @hopfer_social_2014
studied substance use, sexual behavior, and mental health status of
urban population in Winnipeg, Canada. The underlying assumption was that
there were different risk profiles, but their number was not known. 
From a collection of indicators, LCA provided evidence that there are four
distinct risk profiles in the Winnipeg area. Another study investigated 
adolescent empathy development and its influence on parent-child conflict 
[@van_lissa_divergence_2015]. The data was collected over six years. 
Using LCGA, adolescents with similar empathy development trajectories 
were grouped into: high-, average-, and low-empathy classes.

@@

<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
# Best Practices

@@

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection. Special attention should be paid to level of
measurement of the indicators. Finite Gaussian mixture models (including
LPA) are only suitable for continuous variables. Indicators with an
ordinal level of measurement are likely to violate the assumption of
within-class normal distributions of mixture models [see
@vermunt_k-means_2011]. Personal experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
indicators are treated as continuous [e.g., @norman_likert_2010] or 
with 5+ categories [@rhemtulla_when_2012], this
certainly does not imply that all methods are. It is certainly unlikely
that such ordinal variables can be treated as a *mixture* of multiple
normal distributions. The problem becomes egregious when the number of
classes estimated equals or exceeds the number of categories; in this
case, each class-specific mean could describe a single response
category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture
models is that observed indicators are normally distributed
[@spurk_latent_2020]. This is incorrect. When the number of classes is
greater than one, Gaussian mixture models assume that the observed indicators 
are a mixture of multiple (multivariate) normal distributions. In our shoe
size example, it can be seen that the population distribution is
comprised of two normal distributions. When examined visually, the
population distribution is evidently bimodal. The Shapiro-Wilk normality
test ($W = 0.971$, $p < 0.05$ ) rejects the null hypothesis that the
sample comes from a normally distributed population. Yet, this is a
prototypical example of a mixed population distribution where LCA can
discover latent groups. If the population distribution were instead
normal, there would be no classes to extract as the whole population
would belong to a single (homogeneous) class.

>>>>

There are several use cases for which LCA methods are suitable.

### Testing theory

Although latent class analysis is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that latent class analysis is similar to confirmatory factor analysis,
but with a categorical latent variable, it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
One challenge is that, unlike confirmatory factor analysis,
LCA does not provide absolute fit indices that can be used to perform inference.
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values.
A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.

### Unsupervised learning

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be utilized as a diagnostic
aid when using several clinical indicators to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental health
[@wu_abuse_2011] problems. The shoe size example in Figure \@ref(fig:shoedens)
is a rudimentary illustration of this use case.
As another example, a recent study identified latent profiles of high versus low suicide risk in military service members and veterans,
based on self-reported suicidal ideation and information about prior attempted methods and severity
[@gromatsky_characteristics_2022].
Applications like these can aid the clinical decision making process,
and help decide whether additional support is warranted,
or what type of intervention is appropriate.

### Measurement model

Another use case relies on LCA's similarity to confirmatory factor analysis.
This approach is useful when class membership is measured imperfectly by several indicators.
It can be used to partial out measurement error and restore most likely class membership,
and to evaluate the reliability of different indicators.
For example, a recent study had employment status 
information collected across register and survey data. There was disagreement 
between the two data sources. While it might be tempting to assume that the
register data is free of error, this has been shown to be false.
The authors used a mixture model to estimate the
"error corrected employment status" by using two indicators of the same question
and identifying random and systematic measurement error in both indicators 
[@pavlopoulos_measuring_2015].

### Dimension reduction

LCA can also be used to reduce the dimensionality of data,
by reducing many variables to a few prototypes.
This is similar to the use of factor analysis for dimension reduction,
with the key distinction that factor analysis merely accounts for linear covariance between indicators, whereas LCA can accommodate more complex patterns.
As this is a pragmatic application of LCA,
the existence of a true categorical latent variable is questionable (cf. the section on Testing theory).
Consequently, the results of this use case should be treated as descriptive and not reified (i.e., treating the found classes as evidence of the existence of a categorical latent variable) or straightforwardly generalized.
A common use of LCA for dimension reduction is in longitudinal research:
Often, one developmental process is summarized using latent class growth analysis (LCGA), and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy [@van_lissa_divergence_2015].
These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
and that the low empathy group had more conflict with parents than the other groups.
Note that both these key findings are non-linear.
Modeling them in a single-group model would be difficult;
using LCGA greatly simplifies model specification and interpretation,
albeit at a cost of some loss of information and, potentially, generalizability.
<!-- Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean. -->
<!-- Prior to this study, ways to categorize ocean microplastics in terms of length, -->
<!-- width and polymer type were rather arbitrary, and assumed that these properties -->
<!-- were independent. Also, there was a lack of information regarding variability  -->
<!-- within these categories. The authors used finite Gaussian mixture modeling to  -->
<!-- build a classification system for ocean microplastic particles based on a set  -->
<!-- of measured characteristics. With regards to particle dimensions, two classes -->
<!-- emerged: one with smaller fragments whose width and height were highly  -->
<!-- correlated, and another consisting of larger fragments with low correlation  -->
<!-- between particle width and height. -->

### Class enumeration

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group,
but do not know how many groups there are, LCA may be an appropriate
technique to answer this question. For example, @hopfer_social_2014
studied substance use, sexual behavior, and mental health status of
urban population in Winnipeg, Canada. The underlying assumption was that
there were different risk profiles, but their number was not known. 
From a collection of indicators, LCA provided evidence that there are four
distinct risk profiles in the Winnipeg area.

@@

<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

# Best Practices

@@

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection. Special attention should be paid to level of
measurement of the indicators. Finite Gaussian mixture models (including
LPA) are only suitable for continuous variables. Indicators with an
ordinal level of measurement are likely to violate the assumption of
within-class normal distributions of mixture models [see
@vermunt_k-means_2011]. Personal experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
indicators are treated as continuous [e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is very unlikely
that an ordinal variable can be modeled as a *mixture* of multiple
normal distributions. The problem becomes egregious when the number of
classes estimated equals or exceeds the number of categories; in this
case, each class-specific mean could describe a single response
category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture
models is that observed indicators are normally distributed
[@spurk_latent_2020]. This is incorrect. When the number of classes is
greater than one, Gaussian mixture models assume that the observed indicators 
are a mixture of multiple (multivariate) normal distributions. In our shoe
size example, it can be seen that the population distribution is
comprised of two normal distributions. When examined visually, the
population distribution is evidently bimodal. The Shapiro-Wilk normality
test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the
sample comes from a normally distributed population. Yet, this is a
prototypical example of a mixed population distribution where LCA can
discover latent groups. If the population distribution were instead
normal, there would be no classes to extract as the whole population
would belong to a single (homogeneous) class.
